A Shen-Fourier turbulent channel flow solver for large-scale simulations

### 2016-12-09 paper.tex
Major commit

\newcommand{\mmo}[1]{({\bf mmo comment:} \emph{#1})}
\bibliographystyle{plain}

@@

\title{Assessment of a Shen-Fourier turbulent channel flow solver for large-scale simulations}
\author[mmo]{Mikael Mortensen}
\ead{mikaem@math.uio.no}
%\date{}							% Activate to display a given date or no date
\address[mmo]{Department of Mathematics, Division of Mechanics, University of Oslo}

\begin{abstract}
A fully (pseudo-)spectral solver for direct numerical simulations of large-scale turbulent channel flows is described. The solver utilizes the Chebyshev base functions suggested by J. Shen [SIAM J. Sci. Comput., 16, 1, 1995], that lead to stable and robust numerics, even at very large scale. New algorithms for direct solution of the linear systems are devised, and we show that the solver is accurate and applicable for very large simulations. Specifically, the solver is shown to reproduce first order statistics of a channel flow at $Re_{\tau}=2000$ by Hoyas and Jim\'{e}nez [Phys. Fluids, 18(1), 2006], which is higher than has been achieved in a channel by a fully spectral method before. 
\end{abstract}
\begin{keyword}
DNS \sep Fourier \sep Chebyshev
\end{keyword}

@@

The most commonly known DNS use simple geometries, because turbulence physics may then most easily be isolated and studied. Isotropic and homogeneous turbulence are usually studied in triply periodic domains, which allows a spectral Fourier decomposition in all three spatial directions. The equations are solved in spectral space and nonlinearities are usually handled with pseudo-spectral techniques. An important feature is that the pressure can be trivially eliminated, which simplifies the Navier-Stokes equations substantially.

Inhomogeneous, wall bounded, flows, successfully studied with DNS are, e.g., the pipe flow, free shear layers and channel flow between two parallel plates (both Couette and Poiseuille). In this paper we will constrain ourselves to the pressure driven Poiseuille channel flow. In a channel there are two periodic directions that can be handled with Fourier expansions, whereas the wall normal direction is non-periodic and requires a different type of discretization. There are many challenges associated with this inhomogeneity not faced by the pure Fourier solvers, but the first problem at hand is the discretization. The first DNS channel solvers typically used a Chebyshev expansion for the wall-normal direction and, as such, were still able to obtain spectral accuracy in all three spatial directions. A Chebyshev-tau technique, that utilize the recurrence relations of the Chebyshev polynomials, is then used to approximate derivatives, and the coefficient matrices that appear (tridiagonal) can be inverted directly and efficiently. A downside to the Chebyshev-tau method is usually quoted \cite{canuto1988} as numerical stability and roundoff errors, caused by the recurrence relation, and severe condition numbers of the coefficient matrices. For Chebyshev-tau methods the condition numbers have been reported to grow with size as $\mathcal{O}(N^8)$, for a discretization using N points in the wall-normal direction. Discouraged by these numbers, all major recent channel flow simulations have found other, non-spectral, ways of discretizing the non-periodic direction.

The largest known channel simulations to date have been performed by Lee and Moser \cite{leemoser15}, where for $Re_{\tau}\approx 5200$ they used a computational box of resolution $[10240 \times 1530 \times 7680]$ for the streamwise, wall-normal and spanwise directions, respectively. Lee and Moser used seventh-order B-splines for the wall-normal direction. Other simulations of similar magnitude are performed by Hoyas and Jim\'{e}nez \cite{hoyas06, hoyas08}, Lozano-Duran and Jim\'{e}nez \cite{Lozano2014} and Bernardini, Pirozzoli and Orlandi \cite{bernardini2014}. Bernardini et al. used second order finite differences throughout. The Jim\'{e}nez group used dealiased Fourier in the two periodic directions and seventh order compact finite differences, with fourth-order consistency and extended spectral-like resolution \cite{Lele92}, for the wall-normal direction. The solver by Jim\'{e}nez' group is reported to switch from Chebyshev to finite differences if the resolution is above a certain threshold \cite{hoyas08} (reached around $Re_{\tau}=1000$). In other words, they attempt to use spectral methods for as large $Re_{\tau}$ as possible. Spectral methods are attractive for their accuracy, but also for their resolution properties, that are superior to those of any finite difference or spline method. As such, it is desirable to develop fully spectral solvers that can be used for large-scale turbulence simulations.

In his seminal papers \cite{Shen94,Shen95}, Jie Shen describe how to make Legendre and Chebyshev basis functions that lead to sparse matrices, susceptible to fast ($\mathcal{O}(N)$) direct solvers. To the authors knowledge, the bases have not been used for large-scale channel flow simulations, and algorithms for the required direct solvers have not, until now, been devised. Shen's bases have been used for the Navier Stokes equations before, though. Bouchon et.al.  \cite{Bouchon01} describe a spectral-Galerkin formulation very similar to the one used in this paper. However, they choose the Legendre basis over the Chebyshev, due to the simpler structure of the matrices and the lack of weight functions for the Legendre basis. Also, the channel simulations performed were rather small, and there is also the issue of fast transforms for the Legendre basis.

In this paper we will describe and assess a channel flow solver based on Shen's Chebyshev basis \cite{Shen95}. We will give a proper description of the theoretical basis in Sec~\ref{sec:prelim}, the discretization of Navier-Stokes equations in Sec~\ref{sec:discretizationNS} and we will here describe necessary algorithms, including a new fast direct solver for the biharmonic problem that arise in Sec~\ref{sec:implementation}. We will finally show in Sec~\ref{sec:verification} that roundoff is not a major issue, and that the spectral-Galerkin method is in deed applicable to large-scale simulations, larger than what has been achieved for fully spectral channel simulations before.

@@

The Navier Stokes equations are discretized using Fourier basis functions for the periodic directions, and a combination of Chebyshev polynomials in the wall normal direction. Three different sets of basis functions and function spaces are relevant for the wall normal direction
\begin{align}
&  \phi_k(x) = T_k(x), & V_{N_x} &= \text{span}\{\phi_k\}_{k=0}^{N_x} \label{eq:Tk}\\
& \D{\phi}_k(x) = T_k(x) - T_{k+2}(x), & \D{V}_{N_x} &= \text{span} \{ \D{\phi}_k\}_{k=0}^{N_x-2} \label{eq:phiD}\\
& \N{\phi}_k(x) = T_k(x) - \frac{2(k+2)}{k+3} T_{k+2}(x) + 
\frac{k+1}{k+3} T_{k+4}, & \N{V}_{N_x} &= \text{span} \{\N{\phi}_k\}_{k=0}^{N_x-4} \label{eq:phiN} 
\end{align}
where $T_k(x)$ is the $k$'th degree Chebyshev polynomial of the first kind. The 
basis functions and function spaces in (\ref{eq:phiD}) and (\ref{eq:phiN}) were 
suggested by Shen, and satisfy, respectively, the boundary conditions 
$\D{\phi}_k(\pm 1) = 0$, $\N{\phi}_k(\pm 1)=0$ and $\N{\phi}'_k(\pm 1)=0$. The 
last two function spaces may alternatively be written as $\D{V}_{N_x} = \{v \in 
V_{N_x}: v(\pm 1)=0 \}$ and $\N{V}_{N_x} = \{v \in V_{N_x}: v(\pm 1) = v'(\pm 
1) = 0 \}$.

@@

proposed by Kim, Moin and Moser (1987). This scheme is developed by taking the 

@@

The convection vector $\bm{\hat{\mathcal{H}}}$, needed to compute Eqs. (\ref{eq:S_hv}) and (\ref{eq:S_hg}), is found by projecting to the Dirichlet vector space $\D{V}_N^3$, which corresponds to transforming the nonlinear cross product over the entire mesh
\begin{equation}
\bm{\hat{\mathcal{H}}}_i = \mathcal{\D{T}}((\bm{u} \times \bm{\omega})_i) \quad \forall \, i=0,1,2. \label{eq:H_hat}
\end{equation}
The curl, $\bm{\omega} = (g, \partial_z u - \partial_x w, \partial_y u - \partial_x v)$, is computed by projecting each term to its appropriate space. To this end the two terms $\partial_x v$ and $\partial_x w$ are projected to $V_N$, whereas the remaining $\partial_y u$ and $\partial_z u$ are projected to $\N{V}_N$. The nonlinear term is also in need of dealiasing. In this work dealiasing has been performed with the 3/2-rule, see Canuto~et al. \cite{canuto1988}.

@@

\section{Implementation}
\label{sec:implementation}
An outline of the solution procedure used, for the numerical method described 
in Sec.~\ref{sec:discretizationNS}, is given in Algorithm \ref{alg:NS}. Note 
that the Adams-Bashforth projection of convection reverts to a first order 
forward Euler on the first time step. For simulations of turbulent flows this 
is insignificant, but for accuracy tests, like the Orr-Sommerfeld instability, 
a proper initialization would need to set both the two first time steps instead.
\begin{algorithm}
	\caption{Solution procedure for Navier Stokes equations.    
	 \label{alg:NS}}
 	\DontPrintSemicolon	
    	Initialize $\bm{u}(\bm{x}, 0)$ and compute 
    	$\hat{g}_{\bm{k}}(0)$ and $\hat{\bm{u}}_{\bm{k}}(0)$ \;
		Compute LU decomposition of $\N{H}(m,n), \D{H}(m,n) 
		\,\forall \, m, n \in K^p$ \;
		Compute LU decomposition of coefficient matrices in (\ref{eq:v00}) and 
		(\ref{eq:w00}) \;
		$t \gets 0$\; 
		${\kappa} \gets 0$ \;
    	Compute nonlinear convection 
    	$\bm{\hat{\mathcal{H}}}^{\kappa-1}(\hat{\bm{u}}_{\bm{k}}(0), 
    	\hat{g}_{\bm{k}}(0))$ \;
    	\While{$t<T$}{
	    	Compute nonlinear convection $\bm{\hat{\mathcal{H}}}^{\kappa}$ \;
	    	{Compute right hand sides of Eqs.~}(\ref{eq:ufin}), 
	    	(\ref{eq:gfin}) \;
		    {Solve Eq.}(\ref{eq:ufin}) for 
		    $\hat{u}^{{\kappa}+1}_{\bm{k}} \, \forall \,\bm{k} \in \N{K}_N$ \;
		    {Solve Eqs.}(\ref{eq:gfin}) for 
		    $\hat{g}^{{\kappa}+1}_{\bm{k}} \, \forall\, \bm{k} \in \D{K}_N$ \;
		    {Solve Eqs.}(\ref{eq:f3}, \ref{eq:g3}) for  
		    $\hat{v}_{\bm{k}}, \hat{w}_{\bm{k}}\, \forall\, \bm{k} 
		    \in \D{K}_N^0$ \;
		    {Solve Eqs.}(\ref{eq:v00}, \ref{eq:w00}) for 
		    $\hat{v}_0^{{\kappa}+1}, \hat{w}_0^{{\kappa}+1}\, 
		    \forall \, l \in \D{K}^x$ \;
		    ${\kappa} \gets {\kappa} + 1$ \;
		    $t \gets t+\triangle t$ \;
		    {Update to new time step} \;
	    }
\end{algorithm}

The major computational cost in Alg. \ref{alg:NS} comes from  computing the 
nonlinear convection and solving for Eqs.~(\ref{eq:ufin}, \ref{eq:gfin}). The 
nonlinear convection term, $\bm{\hat{\mathcal{H}}}$, is computed using a 
pseudo-spectral method with the 3/2-rule for dealiasing in the periodic 
directions, see \cite{canuto1988}. The 
transforms are carried out using fast Fourier and fast Chebyshev transforms 
that are all of order $\mathcal{O}(N_x\log N_x)$ per one-dimensional transform of size 
$N_x$. The Shen transforms make use of the Chebyshev transforms that are using 
discrete cosine transforms. The forward scalar products and transforms are implemented as shown 
in Algorithm \ref{alg:fst} of the Appendix. A complete transform is obtained through a scalar 
product followed by solving the linear system involving the mass 
matrix, just like shown in Eq. (\ref{eq:fstB}) for space $\N{V}_N$ and similar 
for $\D{V}_N$ and $V_N$. Since the mass matrices are diagonal, tridiagonal and 
pentadiagonal for $V_N, \D{V}_N$ and $\N{V}_N$ respectively, solving the linear 
system is $\mathcal{O}(N_x)$ and the complete transform still scales like $\mathcal{O}(N_x\log N_x)$.

@@

\SetKwProg{Fn}{Function}{}{}
\SetKwFunction{reduceAB}{ReduceAB}
\begin{algorithm}
	\DontPrintSemicolon
    \caption{Recursive formula to compute $\{a_k\}_{k=0}^{N_x-4}, 
	\{b_k\}_{k=0}^{N_x-4}$ in Eq. (\ref{eq:ab}). Note that the lower triangular 
	matrix $\{L_{kj}\}_{k,j=0}^{N_x-4}$ contains only two nonzero diagonals and 
	the parameters $p_k$ and $r_k$ are given in Eqs. (\ref{eq:pk}) and 
	(\ref{eq:rk}) respectively. \label{alg:ab}}
		\Fn{\reduceAB{$ \{\N{L}_{kj}\}_{k,j=0}^{N_x-4}, \,\xi_0$}}{
		\For{$k=0,1$}{
		     $a_k \gets p_k$ \;
		     $b_k \gets r_k$ \;
		}
		\For{$k=2,3$}{
		    $a_k \gets p_k - \N{L}_{k,k-2}\,a_{k-2}$ \;
		    $b_k \gets r_k - \N{L}_{k, k-2}\,b_{k-2}$ \;
		}
		\For{$k=4, 5, \ldots, N_x-4$}{
     		$a_k \gets p_k - \N{L}_{k,k-2}\,a_{k-2} - \N{L}_{k, k-4} \,a_{k-4}$ 
     		\;
	    	$b_k \gets r_k - \N{L}_{k, k-2}\,b_{k-2}- \N{L}_{k, k-4} \,b_{k-4}$ 
	    	\;
		}
		\Return $\{a_k\}_{k=0}^{N_x-4}, \{b_k\}_{k=0}^{N_x-4}$ \;
    }
\end{algorithm}

\SetKwFunction{LUSolveBiharmonic}{LUSolveBiharmonic}
\begin{algorithm}
    \DontPrintSemicolon
	\caption{Solve biharmonic Eq. (\ref{eq:ufin}) with pre-computed 
	$\{\N{U}_{kj}\}_{k,j=0}^{N}, 
		\{\N{L}_{kj}\}_{k,j=0}^{N}$ 
	matrices, as well as the $\{ a_k\}_{k=0}^N$ and $\{ b_k\}_{k=0}^N$ vectors. 
	The parameters $q_j$ and $s_j$ are given in Eqs. (\ref{eq:pk}) and 
	(\ref{eq:rk}) respectively. 	\label{alg:SolveBiharmonic}}
		\Fn{\LUSolveBiharmonic{$\{\N{U}_{kj}\}_{k,j=0}^{N}, 
                \{\N{L}_{kj}\}_{k,j=0}^{N}, 
                \{f_k\}_{k=0}^{N}, \{a_k\}_{k=0}^{N}, \{b_k\}_{k=0}^{N}, 
                \xi_0$}}{
          \SetKwInOut{Output}{output}
          \Output{$\{ \hat{u}_k\}_{k=0}^N$}
		  \tcp*[l]{Solve $\N{L}_{kj}y_j=f_k$ by forward elimination}
		  $y_0 \gets f_0$ \;
		  $y_1 \gets f_1$ \;
		  $y_2 \gets f_2 - \N{L}_{20}y_{0}$ \;
		  $y_3 \gets f_3 - \N{L}_{31}y_{1}$ \;
		  \For{$k = 4$ \textbf{to} $N$}{
		  $y_k \gets f_k - \N{L}_{k,k-2}y_{k-2} - \N{L}_{k, k-4}y_{k-4}$ \;
		  }
		  \tcp*[l]{Solve $\N{U}_{kj}u_j=y_k$ with back 
              substitution}                    
		  $\hat{u}_N \gets y_N / \N{U}_{NN}$ \;
	      $\hat{u}_{N-1} \gets y_{N-1} / \N{U}_{N-1,N-1}$ \;
	      $\hat{u}_{N-2} \gets \left(y_{N-2} - \N{U}_{N-2, 
	      N}y_N\right)/\N{U}_{N-2,N-2}$ \;
	      $\hat{u}_{N-3} \gets \left(y_{N-3} - \N{U}_{N-3, 
	      N-1}y_{N-1}\right)/\N{U}_{N-3,N-3}$ \;

	      float $q^o = q^e = s^o = s^e = 0.0$ \;
		  \For{$k = N-4$ \textbf{to} $0$ \textbf{step} $-1$}{
		  $\hat{u}_k \gets y_k - \N{U}_{k, k+2} \hat{u}_{k+2} - \N{U}_{k, k+4} 
		  \hat{u}_{k+4} $ \;
		  \If{$k < N-5$}{
		  $j \gets k+6$ \;
		  \If{$k$ is odd}{
		    $q^o \gets q^o + \hat{u}_{j}q_{j}$ \;
		    $s^o \gets s^o + \hat{u}_{j}s_{j}$ \;
		    $\hat{u}_k \gets \hat{u}_k - \xi_0q^o a_k - \xi_0s^o b_k $ \;
        }
		  \Else{
		    $q^e \gets q^e + \hat{u}_{j}q_{j}$ \;
		    $s^e \gets s^e + \hat{u}_{j}s_{j}$ \;	  
		    $\hat{u}_k \gets \hat{u}_k - \xi_0q^e a_k - \xi_0s^e b_k $ \;
		}
		}
		  $\hat{u}_k \gets  \hat{u}_k / \N{U}_{kk}$ \;
		  }
		  \Return $\{\hat{u}_k\}_{k=0}^{N}$ \;  
		}
\end{algorithm}

>>>>

\SetKwProg{Fn}{Function}{}{}
\DontPrintSemicolon
\SetKwInOut{Output}{output}
\SetKwInOut{Input}{input}
\SetKw{KwStep}{ step }
\newcommand{\mmo}[1]{({\bf mmo comment:} \emph{#1})}
%\bliographystyle{plain}
\bibliographystyle{model1-num-names}

@@

\title{A Shen-Fourier turbulent channel flow solver for large-scale simulations}
\author[mmo]{Mikael Mortensen}
\ead{mikaem@math.uio.no}
%\date{}							% Activate to display a given date or no date
\address[mmo]{Department of Mathematics, Division of Mechanics, University of Oslo}

\begin{abstract}
A fully (pseudo-)spectral solver for direct numerical simulations of large-scale turbulent channel flows is described. The solver utilizes the Chebyshev base functions suggested by J. Shen [SIAM J. Sci. Comput., 16, 1, 1995], that lead to stable and robust numerics, even at very large scale. New algorithms for direct solution of the linear systems are devised, and we show that the solver is accurate and applicable for high Reynolds numbers and very large simulations. Specifically, the solver is shown to reproduce first order statistics of a channel flow at $Re_{\tau}=2000$ by Hoyas and Jim\'{e}nez [Phys. Fluids, 18(1), 2006], which is higher than has been achieved in a channel by a fully spectral method before. 
\end{abstract}
\begin{keyword}
DNS \sep Fourier \sep Chebyshev
\end{keyword}

@@

The most commonly known DNS use simple geometries, because turbulence physics may then most easily be isolated and studied. Isotropic and homogeneous turbulence are usually studied in triply periodic domains, which allows a spectral Fourier decomposition in all three spatial directions. The equations are most often solved in spectral space and nonlinearities are usually handled with pseudo-spectral techniques. An important feature is that the pressure can be trivially eliminated, which simplifies the Navier-Stokes equations substantially.

Inhomogeneous, wall bounded, flows, successfully studied with DNS are, e.g., the pipe flow, free shear layers and channel flow between two parallel plates (both Couette and Poiseuille). In this paper we will constrain ourselves to the pressure driven Poiseuille channel flow. In a channel there are two periodic directions that can be handled with Fourier expansions, whereas the wall normal direction is non-periodic and requires a different type of discretization. There are many challenges associated with this inhomogeneity not faced by the pure Fourier solvers, but the first problem at hand is the discretization. The first DNS channel solvers typically used a Chebyshev expansion for the wall-normal direction and, as such, were still able to obtain spectral accuracy in all three spatial directions. A Chebyshev-tau technique, that utilize the recurrence relations of the Chebyshev polynomials, was frequently  used to approximate derivatives, and the coefficient matrices that appeared (tridiagonal) could then be inverted directly and efficiently. A downside to the Chebyshev-tau method is usually quoted \cite{canuto1988} as numerical stability and roundoff errors, caused by the recurrence relation, and severe condition numbers of the coefficient matrices. For Chebyshev-tau methods the condition numbers have been reported to grow with size as $\mathcal{O}(N^8)$, for a discretization using $N$ points in the wall-normal direction. Discouraged by these numbers, all major recent channel flow simulations have found other, non-spectral, ways of discretizing the non-periodic direction.

The largest known channel simulations to date have been performed by Lee and Moser \cite{leemoser15}, where for $Re_{\tau}\approx 5200$ they used a computational box of resolution $[10240 \times 1530 \times 7680]$ for the streamwise, wall-normal and spanwise directions, respectively. Lee and Moser used seventh-order B-splines for the wall-normal direction. Other simulations of similar magnitude have been performed by Hoyas and Jim\'{e}nez \cite{hoyas06, hoyas08}, Lozano-Duran and Jim\'{e}nez \cite{Lozano2014} and Bernardini, Pirozzoli and Orlandi \cite{bernardini2014}. Bernardini et al. used second order finite differences throughout. The Jim\'{e}nez group used dealiased Fourier in the two periodic directions and seventh order compact finite differences, with fourth-order consistency and extended spectral-like resolution \cite{Lele92}, for the wall-normal direction. The solver by Jim\'{e}nez' group is reported to switch from Chebyshev to finite differences if the resolution is above a certain threshold \cite{hoyas08} (reached around $Re_{\tau}=1000$). In other words, they attempt to use spectral methods for as large $Re_{\tau}$ as possible. Spectral methods are attractive for their accuracy, but also for their resolution properties, that are superior to those of any finite difference or spline method. As such, it is desirable to develop fully spectral solvers that can be used for large-scale turbulence simulations.

In his seminal papers \cite{Shen94,Shen95}, Jie Shen describe how to construct Legendre and Chebyshev basis functions that lead to sparse matrices, susceptible to fast ($\mathcal{O}(N)$) direct solvers. To the authors knowledge, the bases have not been used for large-scale channel flow simulations, and algorithms for the required direct solvers have not, until now, been devised. Shen's bases have been used for the Navier Stokes equations before, though. Bouchon et.al.  \cite{Bouchon01} describe a spectral-Galerkin formulation very similar to the one used in this paper. However, they choose the Legendre basis over the Chebyshev, neglecting the higher cost of transforms and arguing instead for the simpler structure of the matrices and the lack of weight functions for the Legendre basis. The cost of the Legendre transforms, though, makes the basis unsuitable for larger scales, and the simulations performed in \cite{Bouchon01} were rather small.

In this paper we will describe and assess a channel flow solver based on Shen's Chebyshev basis \cite{Shen95}. We will give a proper description of the theoretical basis in Sec~\ref{sec:prelim}, the discretization of Navier-Stokes equations in Sec~\ref{sec:discretizationNS} and we will describe necessary algorithms, including a new fast direct solver for the biharmonic problem that arise, in Sec~\ref{sec:implementation}. We will finally show in Sec~\ref{sec:verification} that roundoff is not a major issue, and that the Shen-Fourier spectral Galerkin method is in deed applicable to large-scale simulations, larger than what has been achieved for fully spectral channel simulations before.

@@

The Navier Stokes equations are discretized using Fourier basis functions for the periodic directions, and a combination of Chebyshev polynomials in the wall normal direction. Three different sets of basis functions and function spaces are relevant for the wall normal direction
\begin{align}
&  \phi_k(x) = T_k(x), & V_{N_x} &= \text{span}\{\phi_k\}_{k=0}^{N_x}, \label{eq:Tk}\\
& \D{\phi}_k(x) = T_k(x) - T_{k+2}(x), & \D{V}_{N_x} &= \text{span} \{ \D{\phi}_k\}_{k=0}^{N_x-2}, \label{eq:phiD}\\
& \N{\phi}_k(x) = T_k(x) - \frac{2(k+2)}{k+3} T_{k+2}(x) + 
\frac{k+1}{k+3} T_{k+4}, & \N{V}_{N_x} &= \text{span} \{\N{\phi}_k\}_{k=0}^{N_x-4}, \label{eq:phiN} 
\end{align}
where $T_k(x)$ is the $k$'th degree Chebyshev polynomial of the first kind. The 
basis functions and function spaces in (\ref{eq:phiD}) and (\ref{eq:phiN}) were 
suggested by Shen, and satisfy, respectively, the boundary conditions 
$\D{\phi}_k(\pm 1) = 0$, $\N{\phi}_k(\pm 1)=0$ and $\N{\phi}'_k(\pm 1)=0$. The 
last two function spaces may alternatively be written as $\D{V}_{N_x} = \{v \in 
V_{N_x}: v(\pm 1)=0 \}$ and $\N{V}_{N_x} = \{v \in V_{N_x}: v(\pm 1) = v'(\pm 
1) = 0 \}$.

@@

proposed by Kim, Moin and Moser \cite{Kim87}. This scheme is developed by taking the 

@@

The convection vector $\bm{\hat{\mathcal{H}}}$, needed to compute Eqs. (\ref{eq:S_hv}) and (\ref{eq:S_hg}), is found by projecting to the Dirichlet vector space $\D{V}_N^3$, which corresponds to transforming the nonlinear cross product over the entire mesh
\begin{equation}
\bm{\hat{\mathcal{H}}}_i = \mathcal{\D{T}}((\bm{u} \times \bm{\omega})_i) \quad \forall \, i=0,1,2. \label{eq:H_hat}
\end{equation}
The curl, $\bm{\omega} = (g, \partial_z u - \partial_x w, \partial_y u - 
\partial_x v)$, is computed by projecting each term to its appropriate space. 
To this end the two terms $\partial_x v$ and $\partial_x w$ are projected to 
$V_N$, whereas the remaining $\partial_y u$ and $\partial_z u$ are projected to 
$\N{V}_N$. Note that the nonlinear term is also generally in need of 
dealiasing, at least for the two periodic directions.

@@

\section{Implementation}
\label{sec:implementation}
An outline of the solution procedure used, for the numerical method described 
in Sec.~\ref{sec:discretizationNS}, is given in Algorithm 
\ref{alg:NS}.\footnote{Note that throughout this paper we are using the 
pseudocode conventions of Kopriva \cite{kopriva09}, with one minor difference. 
A statement like $\{w\}_{k=0}^{N} \gets \{u\}_{k=0}^N$ will here be used to 
represent a 
vectorization, where components $u_k$, for $k=0,1,\ldots,N$, are copied from 
$u_k$ to $w_k$. Similar conventions apply to matrices, e.g., 
$\{U_{1,k}\}_{k=0}^N = \{V_{2,k}\}_{k=0}^N$ can be used for copying row 2 of 
$V$ to row 1 of $U$.} Note 
that the Adams-Bashforth projection of convection reverts to a first order 
forward Euler on the first time step. For simulations of turbulent flows this 
is insignificant, but for accuracy tests, like the Orr-Sommerfeld instability, 
a proper initialization would need to set both the two first time steps.
\begin{algorithm}
	\caption{Solution procedure for Navier Stokes equations.    
	 \label{alg:NS}}
 	\DontPrintSemicolon	
    	Initialize $\bm{u}(\bm{x}, 0)$ and compute 
    	$\hat{g}_{\bm{k}}(0)$ and $\hat{\bm{u}}_{\bm{k}}(0)$ \;
        Set parameters (e.g., mesh and viscosity) and end time $T$ \;
		Compute LU decomposition of $\N{H}(m,n), \D{H}(m,n) 
		\,\forall \, m, n \in K^p$ \;
		Compute LU decomposition of coefficient matrices in (\ref{eq:v00}) and 
		(\ref{eq:w00}) \;
		$t \gets 0$\; 
		${\kappa} \gets 0$ \;
    	Compute nonlinear convection 
    	$\bm{\hat{\mathcal{H}}}^{\kappa-1}(\hat{\bm{u}}_{\bm{k}}(0), 
    	\hat{g}_{\bm{k}}(0))$ \;
    	\While{$t<T$}{
	    	Compute nonlinear convection $\bm{\hat{\mathcal{H}}}^{\kappa}$ \;
	    	{Compute right hand sides of Eqs.~}(\ref{eq:ufin}), 
	    	(\ref{eq:gfin}) \;
		    {Solve Eq.}(\ref{eq:ufin}) for 
		    $\hat{u}^{{\kappa}+1}_{\bm{k}} \, \forall \,\bm{k} \in \N{K}_N$ \;
		    {Solve Eqs.}(\ref{eq:gfin}) for 
		    $\hat{g}^{{\kappa}+1}_{\bm{k}} \, \forall\, \bm{k} \in \D{K}_N$ \;
		    {Solve Eqs.}(\ref{eq:f3}, \ref{eq:g3}) for  
		    $\hat{v}_{\bm{k}}, \hat{w}_{\bm{k}}\, \forall\, \bm{k} 
		    \in \D{K}_N^0$ \;
		    {Solve Eqs.}(\ref{eq:v00}, \ref{eq:w00}) for 
		    $\hat{v}_0^{{\kappa}+1}, \hat{w}_0^{{\kappa}+1}\, 
		    \forall \, l \in \D{K}^x$ \;
		    ${\kappa} \gets {\kappa} + 1$ \;
		    $t \gets t+\triangle t$ \;
		    {Update to new time step} \;
	    }
\end{algorithm}

The major computational cost in Alg. \ref{alg:NS} comes from  computing the 
nonlinear convection and solving for Eqs.~(\ref{eq:ufin}, \ref{eq:gfin}). The 
nonlinear convection term, $\bm{\hat{\mathcal{H}}}$, is computed using a 
pseudo-spectral method, and dealiasing in the periodic directions can be either 
the 3/2-rule \cite{canuto1988} or the 2/3-rule \cite{Orzag71}. The 
transforms are carried out using fast Fourier and fast Chebyshev transforms 
that are all of order $\mathcal{O}(N_x\log N_x)$ per one-dimensional transform 
of size $N_x$. The Shen transforms make use of the Chebyshev transforms that in 
turn are using 
discrete cosine transforms. The forward scalar products and transforms are implemented as shown 
in Algorithm \ref{alg:fst} of the Appendix. A complete transform is obtained through a scalar 
product followed by solving the linear system involving the mass 
matrix, just like shown in Eq. (\ref{eq:fstB}) for space $\N{V}_N$ and similar 
for $\D{V}_N$ and $V_N$. Since the mass matrices are diagonal, tridiagonal and 
pentadiagonal for $V_N, \D{V}_N$ and $\N{V}_N$ respectively, solving the linear 
system is $\mathcal{O}(N_x)$ and the complete transform still scales like $\mathcal{O}(N_x\log N_x)$.

@@

\SetKwFunction{LUbiharmonic}{LUbiharmonic}
\begin{algorithm}
    \caption{LU decomposition of biharmonic operator $\N{H}$. Current algorithm 
        is using dense storage of $\N{U}_{kj}$, but no more than the 3 items 
        per row need 
        to be stored after the row is no longer needed in the for-loop below 
        (as the loop is traversed to row $k$, $\N{U}_{m, j}$ is no longer 
        needed for $m<k-4$). 
        \label{alg:lu_biharmonic}}
        \Fn{\LUbiharmonic{$\{\N{H}_{kj}\}_{k, j = 0}^{N}$}}{
        \Output{$\{\N{L}_{kj}\}_{k, j = 0}^{N}$, $\{\N{U}_{kj}\}_{k, j = 
        0}^{N}$}
        integer $M_e \gets \lfloor N/2 \rfloor$ \;
        integer $M_o \gets \lfloor (N-1)/2 \rfloor$  \;
        $\{\hat{U}_{0,2j}\}_{j=0}^{M_e} \gets \{\hat{H}_{0,2j}\}_{j=0}^{M_e}$ \;
        $\{\hat{U}_{1,2j+1}\}_{j=0}^{M_o} \gets 
        \{\hat{H}_{1,2j+1}\}_{j=0}^{M_o}$ \;
        $\N{L}_{20} \gets \N{H}_{20}/\N{U}_{00}$ \;
        $\N{L}_{31} \gets \N{H}_{31}/\N{U}_{11}$ \;
        $\{\hat{U}_{2,2j}\}_{j=1}^{M_e} \gets \{\N{H}_{2,2j}\}_{j=1}^{M_e} - 
        \N{L}_{20} \{\N{U}_{0,2j}\}_{j=1}^{M_e}$ \;
        $\{\hat{U}_{3,2j+1}\}_{j=1}^{M_o} \gets \{\N{H}_{3,2j+1}\}_{j=1}^{M_o} 
        - \N{L}_{31} \{\N{U}_{1,2j+1}\}_{j=1}^{M_o}$ \;

        %\State $\N{U}_{2j} = \N{H}_{2j} - \N{L}_{20} \N{U}_{0j} \quad 
        %\forall \, j \in 2, 4, \ldots, N$
        %\State $\N{U}_{3j} = \N{H}_{3j} - \N{L}_{31} \N{U}_{1j} \quad 
        %\forall \, j \in 
        %3, 5, \ldots, N$
        \For{$k = 4$ \KwTo $N$}{
        integer $M \gets \lfloor \frac{N-k}{2}\rfloor$ \;
        $\N{L}_{k,k-4} \gets \N{H}_{k, k-4} / \N{U}_{k-4, k-4}$ \;
        $\{\N{U}_{k,k+2j}\}_{j=-1}^{M} \gets \{\N{H}_{k,k+2j}\}_{j=-1}^{M} - 
        \N{L}_{k, k-4} 
        \{\N{U}_{k-4, k+2j}\}_{j=-1}^{M}$ \;
        %\State $\N{U}_{k,j} = \N{H}_{k,j} - \N{L}_{k, k-4}  \N{U}_{k-4, j} 
        %\quad \forall j = k-2, k, \ldots, N$
        $\N{L}_{k,k-2} \gets {\N{U}}_{k,k-2}/{\N{U}}_{k-2, k-2}$ \;
        $\{\N{U}_{k,k+2j}\}_{j=0}^{M} \gets \{{\N{U}}_{k,k+2j}\}_{j=0}^{M} - 
        \N{L}_{k, k-2} \{\N{U}_{k-2,k+2j}\}_{j=0}^{M}$ \;
        %\State $\N{U}_{kj} = {\N{U}}_{kj} - \N{L}_{k, k-2} \N{U}_{k-2, j} 
        %\quad \forall j = k, k+2, \ldots, N$
        }
        \Return $\{\N{L}_{kj}\}_{k, j = 0}^{N}$, $\{\N{U}_{kj}\}_{k, j = 0}^{N}$
    }
\end{algorithm}
